---
title: "MAGeCK Comparison Report"
output: html_notebook
---

<!-- 

This is a template file for R markdown used in MAGeCK 
 
-->

Author: Wei Li, weililab.org


## Parameters

comparison_name is the prefix of your output file, defined by the "-n" parameter in your "mageck test" command. The system will look for the following files to generate this report:

* comparison_name.sgrna_summary.txt
* comparison_name.gene_summary.txt
* comparison_name.log


```{r}
# define the comparison_name here; for example,
# comparison_name='demo'
comparison_name='dmso_test'
```

FDR cutoff is used to draw a boundary line in RRA or p value plots. Set it to -1 to disable the boundary line.
```{r}
fdrcutoff=0.05
# fdrcutoff=-1 # disable FDR cutoff line
```


## Preprocessing


```{r echo=FALSE}


####  function definition #

library(knitr)

genreporttable<-function(cpname,genetb){
  comparisons=rep(cpname,2)
  #genetb=read.table(genesummaryfile,header = T,as.is = T)
  ngenes=rep(nrow(genetb),2)
  direction=c('neg','pos')
  fdr1=c(sum(genetb[,5]<0.01),sum(genetb[,11]<0.01))
  fdr5=c(sum(genetb[,5]<0.05),sum(genetb[,11]<0.05))
  fdr25=c(sum(genetb[,5]<0.05),sum(genetb[,11]<0.05))
  xtb=data.frame(Comparison=comparisons,Genes=ngenes,Selection=direction,FDR1=fdr1,FDR5=fdr5,FDR25=fdr25);
  colnames(xtb)=c("Comparison","Genes","Selection","FDR1%","FDR5%","FDR25%");
  return (xtb);
}


colors=c( "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",  "#A65628", "#F781BF",
          "#999999", "#66C2A5", "#FC8D62", "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", 
          "#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", "#B3DE69", "#FCCDE5",
          "#D9D9D9", "#BC80BD", "#CCEBC5", "#FFED6F")


plotrankedvalues<-function(val, tglist, cutoffline=-1, ...){
  
  plot(val,log='y',ylim=c(max(val),min(val)),type='l',lwd=2, ...)
  if(length(tglist)>0){
    for(i in 1:length(tglist)){
      targetgene=tglist[i];
      tx=which(names(val)==targetgene);ty=val[targetgene];
      points(tx,ty,col=colors[(i %% length(colors)) ],cex=2,pch=20)
      # text(tx+50,ty,targetgene,col=colors[i])
    }
    legend('topright',tglist,pch=20,pt.cex = 2,cex=1,col=colors)
  }
  if(cutoffline>0){
    lines(c(cutoffline,cutoffline),c(min(val)*1e-5,max(val)*10000),lty=2)
  }
}



plotrandvalues<-function(val,targetgenelist,cutoffline=-1, ...){
  # choose the one with the best distance distribution
  
  mindiffvalue=0;
  randval=val;
  for(i in 1:20){
    randval0=sample(val)
    vindex=sort(which(names(randval0) %in% targetgenelist))
    if(max(vindex)>0.9*length(val)){
      # print('pass...')
      next;
    }
    mindiffind=min(diff(vindex));
    if (mindiffind > mindiffvalue){
      mindiffvalue=mindiffind;
      randval=randval0;
      # print(paste('Diff: ',mindiffvalue))
    }
  }
  plot(randval,log='y',ylim=c(max(randval),min(randval)),pch=20,col='grey', ...)
  
  if(length(targetgenelist)>0){
    for(i in 1:length(targetgenelist)){
      targetgene=targetgenelist[i];
      tx=which(names(randval)==targetgene);ty=randval[targetgene];
      points(tx,ty,col=colors[(i %% length(colors)) ],cex=2,pch=20)
      text(tx+length(randval)/100,ty,targetgene,col=colors[i],adj=1,pos=4)
    }
  }
  if(cutoffline>0){
    lines(c(-100,length(randval)*2),c(cutoffline,cutoffline),lty=2)
  }
  
}

getcomparisonsfromlogfile<-function(logfile){
  
  logline=readLines(file(logfile,"r"),n = 1)
  #print(logline)
  logbk=strsplit(logline,' ')[[1]]
  ctrl_sample=c()
  treat_sample=c()
  for(i_x in 1:length(logbk)){
    if(logbk[i_x]=='-c' | logbk[i_x]=='--control'){
      ctrl_sample=strsplit(logbk[i_x+1],',')[[1]]
    }
     if(logbk[i_x]=='-t' | logbk[i_x]=='--treatment'){
      treat_sample=strsplit(logbk[i_x+1],',')[[1]]
    }
  }
  #close(logff)
  return (list(ctrl_sample,treat_sample))
}

plotindvidualsgrnas<-function(sg_table,targetgene,collabel){
  
  xsg_c=sg_table[which(sg_table[,'Gene']==targetgene),c('control_count','treatment_count')]
  rownames(xsg_c)=sg_table[which(sg_table[,'Gene']==targetgene),'sgrna']
  
  xsg_c2=paste(xsg_c[,1],xsg_c[,2],sep='/')
  
  xsg_c3=t(data.frame(lapply(strsplit(xsg_c2,'/'),as.numeric)))
  rownames(xsg_c3)=rownames(xsg_c)
  
  targetmat=xsg_c3
  
  targetmatvec=unlist(targetmat)+1
  yrange=range(targetmatvec[targetmatvec>0]);
  # yrange[1]=1; # set the minimum value to 1
  for(i in 1:nrow(targetmat)){
    vali=targetmat[i,]+1;
    if(i==1){
      plot(1:length(vali),vali,type='b',las=1,pch=20,main=paste('sgRNAs in',targetgene),ylab='Read counts',xlab='Samples',xlim=c(0.7,length(vali)+0.3),ylim = yrange,col=colors[(i %% length(colors))],xaxt='n',log='y')
      if(length(collabel)==length(vali)){
         axis(1,at=1:length(vali),labels=(collabel),las=2)
      }else{
        warning('Cannot extract sample labels.')
      }

      # lines(0:100,rep(1,101),col='black');
    }else{
      lines(1:length(vali),vali,type='b',pch=20,col=colors[(i %% length(colors))])
    } 
  }
  

}



plot_rra_or_p_scores<-function(gstable,startindex,fdrcolumn,target_list,...){
  pvec=gstable[,startindex]
  names(pvec)=gstable[,'id']
  pvec=sort(pvec);
  
  ncutoff=ifelse(fdrcutoff>0,sum(gstable[,fdrcolumn]<fdrcutoff),-1)
  
  vcutoff=ifelse(fdrcutoff>0 & ncutoff>0,max(pvec[1:ncutoff]),-1)
  
  plotrankedvalues(pvec,target_list,cutoffline = ncutoff,...)
  
  plotrandvalues(pvec,target_list,cutoffline = vcutoff,...)

}
```


```{r echo=FALSE}

gene_summary_file=paste(comparison_name,'.gene_summary.txt',sep='')
sgrna_summary_file=paste(comparison_name,'.sgrna_summary.txt',sep='')
log_file=paste(comparison_name,'.log',sep='')

```


Reading input files. If any of these files are problematic, an error message will be shown below.

```{r}
gstable=read.table(gene_summary_file,header = T,as.is = T,na.strings='')
sg_table=read.table(sgrna_summary_file,header = T,as.is = T,na.strings='')
comp_samples=getcomparisonsfromlogfile(log_file)
collabel=c(comp_samples[[1]],comp_samples[[2]])
```

## Summary




The samples used in the comparison is indicated as follows.


```{r echo=FALSE,results='asis'}
sampletable=data.frame(Sample=c(comp_samples[[1]],comp_samples[[2]]),Type=c(rep('control',length(comp_samples[[1]])),rep('treatment',length(comp_samples[[2]]))))
kable(sampletable,caption='Sample summary')
```

The statistics of comparisons is as indicated in the following table. 

```{r echo=FALSE,results='asis'}

cptable=genreporttable(comparison_name,gstable)

kable(cptable,caption='Comparion summary')
```


The meanings of the columns are as follows.

* Comparison: The label for comparisons;
* Genes: The number of genes in the library;
* Selection: The direction of selection, either positive selection or negative selection;
* FDR1%: The number of genes with FDR < 1%;
* FDR5%: The number of genes with FDR < 5%;
* FDR25%: The number of genes with FDR < 25%;

## Plotting invidivual genes in negative selection

The following figures show:

* Individual sgRNA read counts of selected genes in selected samples; 
* The distribution of RRA scores and p values of all genes; and
* The RRA scores and p values of selected genes.

The following genes are used in the plot (change it as you like). By default, it is the top 5 genes in negatively selected genes.

```{r}
targetgenelist_neg=gstable[gstable[,6]<5,1]

# or, directly specify the genes to be plotted
#targetgenelist_neg=c("ACTR8","ACIN1")

# display genes used in the plot
print(targetgenelist_neg)

```

The following figure plots the distribution of RRA scores across these genes. Dotted lines represent the FDR cutoff line defined by the "fdrcutoff" value in the "Paramters" section.

```{r echo=FALSE}

startindex=3
fdrcolumn=5

plot_rra_or_p_scores(gstable,startindex,fdrcolumn,target_list = targetgenelist_neg, xlab='Genes',ylab='RRAscore',main=paste('Distribution of RRA scores in',comparison_name))


```


The following figure plots the distribution of p values in these genes. Dotted lines represent the FDR cutoff line defined by the "fdrcutoff" value in the "Paramters" section.

```{r echo=FALSE}


startindex=4
fdrcolumn=5

plot_rra_or_p_scores(gstable,startindex,fdrcolumn,target_list = targetgenelist_neg, xlab='Genes',ylab='p value',main=paste('Distribution of p values in',comparison_name))


```



## Plotting invidivual genes in positive selection



The following genes are used in the plot (change it as you like). By default, it is the top 5 genes in negatively selected genes.

```{r}

targetgenelist_pos=gstable[gstable[,12]<5,1]

# or, directly specify the genes to be plotted
#targetgenelist_pos=c("ACTR8","ACIN1")

# display genes used in the plot
print(targetgenelist_pos)

```

The following figure plots the distribution of RRA scores across these genes. Dotted lines represent the FDR cutoff line defined by the "fdrcutoff" value in the "Paramters" section.

```{r echo=FALSE}

startindex=9

fdrcolumn=11

plot_rra_or_p_scores(gstable,startindex,fdrcolumn,target_list = targetgenelist_pos, xlab='Genes',ylab='RRAscore',main=paste('Distribution of RRA scores in',comparison_name))


```



The following figure plots the distribution of p values in these genes. Dotted lines represent the FDR cutoff line defined by the "fdrcutoff" value in the "Paramters" section.

```{r echo=FALSE}
startindex=10

fdrcolumn=11

plot_rra_or_p_scores(gstable,startindex,fdrcolumn,target_list = targetgenelist_pos, xlab='Genes',ylab='p value',main=paste('Distribution of p values in',comparison_name))

```

# sgRNA changes

The following figures show the distribution of sgRNA read counts (normalized) of selected genes in selected samples.


```{r}

for(target_gene in c(targetgenelist_neg,targetgenelist_pos)){
  plotindvidualsgrnas(sg_table,target_gene,collabel)
}

```


